\Wxxsir methods to estimate $\rr$ in the SIR model

### 2018-07-09 paper/Version15/draft_v15.tex
Added in Bill edits up to Section 3

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways,$\rr$ summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

@@

Even when limiting $\rr$ estimates to those from SIR models, estimation is still difficult.  Problems include mathematical versus statistical methods (i.e. solving for $\rr$ versus estimating $\rr$), the number of observations used to estimate $\rr$ (i.e. time), boundary cases of infection and recovery rates, population size, initial SI ratio, and restrictive assumptions placed upon the noise.  These are all issues for producing point estimates, to say nothing of confidence or credible intervals (CI).

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate to changes in the model parameters.  For example, the SIR model is special in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of susceptibles is non-decreasing.  Logically, models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the non-symmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.  Without assuming Gaussian noise, it becomes much more difficult to guarantee properties such as unbiased and consistent estimates of $\rr$.

These difficulties in estimating $\rr$ have led a number of researchers to analyze the sensitivity of $\rr$ with respect to small perturbations in parameters like $\beta$ and $\gamma$ \citep{lash2003,epstein2007agent,capaldi2012}.  Even this is a difficult task. Since there is no analytic solution to the SIR model, estimating the set of derivatives with respect to $\beta$ and $\gamma$ must be done numerically, which can substantially increase computation time.  We should note that this sensitivity analysis is done before even adding any distributional assumptions on how randomness enters the model.

Below in Table \ref{tab:r0-real-ex}, we provide examples of estimates of $\rr$ for specific diseases from publications over the past three decades, spanning various locations over the world.  We display estimates for HIV, Zika (ZIK-V), Ebola (EVD), seasonal influenza, H1N1 influenza, and measles.  These estimates of $\rr$ are not limited to those of SIR model, although most results come from this framework.  Of these, Zika seems to have the largest estimated values of $\rr$, followed by Ebola, H1N1 influenza, seasonal influenza, and HIV.  We see that African locations seem to have larger estimates of $\rr$ compared to the rest of the world, regardless of the disease.  Some of the accompanying intervals (not necessarily CIs) are extremely large such as the case of HIV in Uganda (Interval: [0.12, 14.17]) whereas some are extremely small such as for Ebola in Guinea (Interval: [1.50, 1.52]).  We see that estimates for H1N1 influenza are generally higher than that of seasonal influenza, which is evidence that H1N1 should have been treated as a more critical concern than the typical seasonal influenza cycle.  The lowest possible estimate for $\rr$ reported is 0.12 for HIV in Uganda, and the largest is 17.00 for measles in England and Wales.  Within diseases, we see that estimates for $\rr$ seem to be similar, although typically do not lie in one another's intervals, suggesting that there is disparity in estimates of $\rr$.

@@

@@

ZIK-V                & Florida                          & 2016                          & 0.16        & 0.13           & 0.19           & \cite{dinh2016}       \\
ZIK-V                & Barranquilla, Colombia           & 2015                          & 2.40         & 3.80            & 5.60            & \cite{towers2016}    \\
ZIK-V                & Colombia                         & 2015                          & 1.42        & 2.56           & 3.83           & \cite{majumder2016}  \\ \hline

@@

EVD                 & Liberia                          & 2014                          & 1.59        & 1.57           & 1.60            & \cite{althaus2014}    \\ \hline
Seasonal Influenza  & USA, France, Australia           & 1972-2002 & 1.30         & 1.20            & 1.40            & \cite{chowell2008}  \\

@@

Measles      & Cirencester, England                 & 1947-1950                          & 13.50   & 13.00           & 14.00            & \cite{fraser2009}     \\ 

@@

  Measles     & Kansas, USA                  & 1918-1921                          & 5.50    & 5.00          & 6.00          & \cite{anderson1992}     \\
  Measles     & Ontario, Canada                  & 1912-1913                         & 11.50        & 11.00           & 12.00            & \cite{anderson1992}     \\
  Measles     & Willesden, England                  & 1912-1913                          & 11.00        & 12.00       & 1.60            & \cite{anderson1992}     \\

@@

@@

The SIR model is represented graphically in Figure \ref{fig::sir}. 
\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center}]
    \node [int,  fill = white!70!blue] (a)              {$X$};
    \node [int,  fill = white!70!blue]           (c) [right of=a] {$Y$};
    \node [int,  fill = white!70!blue] (e) [right of=c] {$Z$};
    \path[->, auto=false] (a) edge node {} (c)
                          (c) edge node {} (e) ;

\end{tikzpicture}
\caption{Depiction of a SIR model where $X=S, Y=I,$ and $Z=R$.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
An outbreak occurs if the rate of change of infectious individuals is positive, $\frac{dY}{dt} > 0$, or equivalently,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \implies -\frac{dX}{dt} - \frac{dZ}{dt} &> 0 \\
  \implies \frac{\beta X Y}{N}  - \gamma Y &> 0 ,\\
\implies  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0\\
\implies   \frac{\beta}{\gamma} &> \frac{N}{X}.
\end{align*}
That is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of initially susceptible individuals is large compared to the total population, $\frac{X}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align*}
  \rr \overset{def}{=} \frac{\beta}{\gamma}.
  \end{align*}
  To incorporate randomness into the model, we add noise, namely,
  \begin{align}\label{eq:sir-noise}
    X_{obs}(t) &= X(t) + \epsilon_{X,t}\\
    Y_{obs}(t) &=  N - X_{obs}(t) -Y_{obs}(t)  \nonumber\\
    Z_{obs}(t) &= Z(t) - \epsilon_{Z,t}. \nonumber
  \end{align}
We are assuming the observations are generated based on the discrete-time,  deterministic ODEs presented in Equation \ref{eq:sir} with time and compartment dependent noise $\epsilon_{X,t}$ and $\epsilon_{Z,t}$.  Since, $N$, the total population, is constant, then $Y_{obs}$ is adjusted accordingly.

In summary, when we discuss estimators for $\rr$ for the SIR model, we mean to say we are forming an estimator of $\rr$ from the given set of data in Eq. \ref{eq:sir-noise},
\begin{align*}
  \textnormal{Data} &= \left \{\left (X_{obs}(t), Y_{obs}(t), Z_{obs}(t) \right ) : t=0, 1, \dots, T\right \}, \\
  \hat{\rr} &= m(\textnormal{Data}),
\end{align*}
where $m$ is a function of the data.

Good properties of an estimator may include (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} \rr \textnormal{ as } T\to \infty,
\end{align*}
where `$\overset{P}{\to}$' stands for convergence in probability or (2) converging to a known probability distribution (convergence in distribution `$\overset{d}{\to}$' \citep{wasserman2004}),
\begin{align*}
\hat{\rr} \overset{d}{\to} F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr < 1 \\
  H_A:\;& \rr \ge 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $\hat{a}, \hat{b} >0$
\begin{align*}
P(\hat{a} \le \rr \le \hat{b}) \le 1 - \alpha.
\end{align*}
This property is particularly desirable, as we can estimate $\hat{a}$ and $\hat{b}$ from the distribution $F$.  At any rate, we would like to be able to estimate $\hat{a}$ and $\hat{b}$ for a given $\alpha$-level.  We could also make similar hypothesis tests such as that the $\rr$ for H1N1 influenza is greater than the reported $\rr$ for seasonal influenza.

@@

\begin{align*}
  \frac{dX}{dt} &= - \frac{\beta XY}{N} \\
  \frac{dE}{dt} &= \frac{\beta XY}{N}  - \mu E\\
  \frac{dY}{dt} &= \mu E - \gamma Y \\
  \frac{dZ}{dt} &= \gamma Y,
\end{align*}
again, with a constant population of $N$ and with known initial values $(X(0), E(0), Y(0), Z(t))$.  The parameter $\mu$ is the rate at which latently exposed individuals become infectious.  As $\frac{1}{\mu} \to 0$, the XEYZ model converges to the XYZ model.  The reproduction number for the XEYZ model, is the same as in the XYZ model, that is $\rr = \frac{\beta}{\gamma}$.

Consider the case when $N=10000, (\beta, \gamma) = (.06, .03)$, $(X(0), E(0), Y(0), Z(t)) = (9500, 0, 500, 0)$ for times $t=1, \dots, 365$.  Let the parameter $\mu$ range between [0.005, 0.250].  This generates data set
$$D_{\mu, \textnormal{XEYZ}} = \left \{(X_{XEYZ}(t), E_{XEYZ}(t), Y_{XEYZ}(t), Z_{XEYZ}(t): t=1, \dots, 365\right \},$$
which is observed by researcher 1.  Now assume that researcher 2 misspecifies the model for this data as a SIR (XYZ) model rather than a SEIR (XEYZ) model.  That is, she observes all non-infectious individuals as susceptibles and generates data set
$$D_{\mu, XYZ} = \left \{X_{XYZ}(t) = X_{XEYZ} + E_{XEYZ}, Y_{XYZ} = Y_{XEYZ}, Z_{XEYZ} = Z_{XYZ} : t = 1, \dots, 365 \right \}.$$
This situation is illustrated in Figure \ref{fig:sir-vs-seir} with $\mu = 0.01$.

@@

Both researchers want to estimate $\rr$ and so fit models to their data sets.  The details of how the models are fit are not so important at the moment (but will be in the later sections), but to fairly compare one model to another assume both researchers use ratio estimator to fit the models.  That is researcher 1 estimates $\rr$ with
\begin{align*}
  (\hat{\beta}_{XEYZ}, \hat{\gamma}_{XEYZ}, \hat{\mu}_{XEYZ} )&=\arg \min_{\beta, \gamma, \mu} \sum_{t} \left [ \left (X_{XEYZ}(t) - X(t)\right )^2  \right . \\
                                                                 &+\left ( E_{XEYZ}(t) - E(t) \right )^2  \\
  &\left . + \left ( Z_{XEYZ}(t) - Z(t) \right )^2\right ]
\end{align*}
Then the ratio estimator (RE) for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XEYZ}= \frac{\hat{\beta}_{XEYZ}}{\hat{\gamma}_{XEYZ}}.
\end{align*}
Researcher 2 estimates $\rr$ with 
\begin{align*}
(\hat{\beta}_{XYZ}, \hat{\gamma}_{XYZ} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{XYZ}(t) - X(t)\right )^2 + \left ( Z_{XYZ}(t) - Z(t) \right )^2 \right ].
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XYZ}= \frac{\hat{\beta}_{XYZ}}{\hat{\gamma}_{XYZ}}.
\end{align*}

We plot the two estimates of $\rr$ as a function of $\mu$ in Figure \ref{fig:r0-alpha}.  We should note we also estimated the standard errors of the estimates of $\rr$, but they do not show up on the figure because they are of an order of magnitude of $10^{-4}$.  Unsurprisingly, under the SEIR model speficiation, $\hat{\rr}=2$ because there is no noise in the data, and the data is generated from an SEIR model.  However, as we see in Figure \ref{fig:r0-alpha} the RE of $\rr$ from the SIR model for small values of $\mu$ can result in double the size of $\rr$!

@@

For those readers not convinced by our small example, we offer this mathematical intuition of the next generation model (NGM) described by \cite{diekmann1990} and \cite{diekmann2009}.

 The NGM is a generalization of any compartment model at the infection-free steady state ($Y(t)=0$ and all derivatives with respect to time sum equal zero). Originally, introduced by \citep{diekmann1990}, \cite{diekmann2009} posit a recipe to find $\rr$ for a wide class of compartment models, including SIR and SEIR. First, define the infected subsystem as ``the equations of the ODE system that describe the production of new infections and changes in state among infected individuals.''  Let $x = (C_1, C_2, \dots, C_m)^T$ where $C_i$ are the different compartments of the infected subsystem.  The steps to find $\rr$ are as follows:

\begin{enumerate}
\item ``Linearize the infected subsystem of nonlinear ODEs about the infection-free steady state''
\item Decompose the linearized infected subsystem into $(T + \Sigma )x$ where $T$ is a $m\times m$ matrix of transmissions and $\Sigma$ is a $m \times m$ matrix of transitions.
\item $\rr$ is the spectral radius (i.e. the dominant eigenvalue) of $-T \Sigma^{-1}$.  
\end{enumerate}

@@

The NGM shows that $\rr$ is the spectral radius of a matrix composed from the transmission of disease from diseased individuals to susceptible individuals and transition of diseased individuals from among the compartments and is indeed a property of the model.

Following the examples presented in \cite{diekmann1990}, for the SIR (XYZ) model, the linearized infection-free sub-system occurs when $Y(t)=0$ and $X(t)=N$ and is
\begin{align*}
\frac{dY}{dt} &= - \beta Y + \gamma .
  \end{align*}
  Then $T= \beta$ (the rate of new infections) and $\Sigma = -\gamma$.  Then $K = \frac{\beta}{\gamma}$ which is also, trivially, the dominant eigenvalue, and hence, $R_0 = \frac{\beta}{\gamma}$.

We can also derive the NGM for the SEIR (XEYZ) model.  We have an infection-free steady state when $E=Y=0$ and set $X=N$.  Then $x = (E, Y)^T$ is the infected subsystem.  The linearized subsystem is
  \begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dE}{dt} &= \beta Y  - \mu E\\
    \frac{dY}{dt} &=  \mu E  - \gamma Y \\
  \end{array}
  \right .
  \end{align*}
  Then $T = \left ( \begin{array}{cc} 0 & \beta \\ 0 & 0  \end{array} \right )$ and $\Sigma = \left ( \begin{array}{cc} -\mu & 0 \\ \mu & - \gamma \end{array} \right )$, and $K =\left ( \begin{array}{cc} 0& \beta / \gamma \\ 0 & 0 \end{array} \right  )$.

  Then the spectral radius of $K$ is $R_0  = \frac{\beta}{\gamma}$, which makes sense as exposed individuals are not contributing to the number of new infections.

  Although the $\rr$ ends up being the same in both the SIR (XYZ) and SEIR (XEYZ) case, the derivations were different because of the addition of the exposed compartment.  In general, more complicated models will result in different derivations of $T$ and $\Sigma$ and hence different derivations of $\rr$.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{obs}(t) - X(t)\right )^2 + \left ( Z_{obs}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the ratio estimator (RE) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

The estimates from this method are often found via a grid search of $\beta$ and $\gamma$ values to minimize the squared error.  The estimates can also be found using more sophisticated optimization algorithms. Since we typically cannot explicitly write down the partial derivatives of $S$, $I$, and $R$ with respect to $\beta$ and $\gamma$, we can only use minimization algorithms which do not rely on an explicit gradient such as Nelder-Mead simplex minimization \citep{nelder-mead1965}.

>>>>

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the initial reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways, $\rr$ summarizes an entire outbreak of a disease; it  is used to assess whether a disease epidemic will occur.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$ \citep{anderson1992}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

@@

Even when limiting $\rr$ estimates to those from SIR models, estimation is still difficult.  Problems include mathematical versus statistical methods (i.e. solving for $\rr$ versus estimating $\rr$), the number of observations used to estimate $\rr$ (i.e. time), boundary cases of infection and recovery rates, population size, initial SI ratio, and restrictive assumptions placed upon the noise.  These are all issues for producing point estimates, to say nothing of issues related to confidence or credible intervals (CI).

@@

Other difficulties in estimating $\rr$ arise from model assumptions on how randomness enters the model and sensitivity of the estimate to changes in the model parameters.  For example, the SIR model is special in that two of its compartments, S and R are monotonic.  The number of susceptibles is non-increasing and the number of susceptibles is non-decreasing.  Logically, models should enforce this monotonicity of the two compartments.  As a consequence, this restricts assumptions on how randomness enters the model.  For example, Gaussian error, a common assumption in statistics and machine learning, becomes highly suspect due to the non-symmetrical nature of the noise imposed by the monotonic $S$ and $R$ compartments.

These difficulties in estimating $\rr$ have led a number of researchers to analyze the sensitivity of $\rr$ with respect to small perturbations in parameters like $\beta$ and $\gamma$, respectively the infection and recovery rate \citep{lash2003,epstein2007agent,capaldi2012}.  Even this is a difficult task. Since there is no analytic solution to the SIR model, estimating the set of derivatives with respect to $\beta$ and $\gamma$ must be done numerically, which can substantially increase computation time.  We should note that this sensitivity analysis is done before even adding any distributional assumptions on how randomness enters the model.

Below in Table \ref{tab:r0-real-ex}, we provide examples of estimates of $\rr$ for specific diseases from publications over the past three decades, spanning various locations over the world.  We display estimates for HIV, Zika (ZIK-V), Ebola (EVD), seasonal influenza, H1N1 influenza, and measles.  These estimates of $\rr$ are not limited to those of SIR model, although most results come from this framework.  Of these, Zika seems to have the largest estimated values of $\rr$, followed by Ebola, H1N1 influenza, seasonal influenza, and HIV.  We see that African locations seem to have larger estimates of $\rr$ compared to the rest of the world, regardless of the disease.  Some of the accompanying intervals (not necessarily CIs) are extremely large such as the case of HIV in Uganda (Interval: [0.12, 14.17]) whereas some are extremely small such as for Ebola in Guinea (Interval: [1.50, 1.52]).  We see that estimates for H1N1 influenza are generally higher than that of seasonal influenza, although we find that in our appliaction the opposite is true.  The lowest possible estimate for $\rr$ reported is 0.12 for HIV in Uganda, and the largest is 17.00 for measles in England and Wales.  Within diseases, we see that estimates for $\rr$ seem to be similar, although typically do not lie in one another's intervals, suggesting that there is disparity in estimates of $\rr$.

@@

ZIK-V                & Colombia - Barrranquilla           & 2015                          & 2.40         & 3.80            & 5.60            & \cite{towers2016}    \\  
ZIK-V                & Colombia                         & 2015                          & 1.42        & 2.56           & 3.83           & \cite{majumder2016}  \\

@@

  ZIK-V                & Florida                          & 2016                          & 0.16        & 0.13           & 0.19           & \cite{dinh2016}       \\ \hline
  EVD                 & Liberia                          & 2014                          & 1.59        & 1.57           & 1.60            & \cite{althaus2014}    \\

@@

\hline
Seasonal Influenza  & Australia, France, USA          & 1972-2002 & 1.30         & 1.20            & 1.40            & \cite{chowell2008}  \\

@@

  Measles     & Canada - Ontario                  & 1912-1913                         & 11.50        & 11.00           & 12.00            & \cite{anderson1992}     \\
  Measles      & England - Cirencester                 & 1947-1950                          & 13.50   & 13.00           & 14.00            & \cite{fraser2009}     \\
  Measles     & England - Willesden                  & 1912-1913                          & 11.00        & 12.00       & 1.60            & \cite{anderson1992}     \\  

@@

@@

  Measles     & USA - Kansas                  & 1918-1921                          & 5.50    & 5.00          & 6.00          & \cite{anderson1992}     \\

@@

The SIR model is represented graphically in Figure \ref{fig::sir}. 
\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center}]
    \node [int,  fill = white!70!blue] (a)              {$X$};
    \node [int,  fill = white!70!blue]           (c) [right of=a] {$Y$};
    \node [int,  fill = white!70!blue] (e) [right of=c] {$Z$};
    \path[->, auto=false] (a) edge node {} (c)
                          (c) edge node {} (e) ;
\end{tikzpicture}
\caption{Depiction of a SIR model where $X=S, Y=I,$ and $Z=R$.  One can only be infected once by the disease once in this model.}\label{fig::sir}
\end{figure}
Since the number of indivudals ($N$) is constant, then
\begin{align*}
\frac{dX}{dt} + \frac{dY}{dt} + \frac{dZ}{dt} \equiv 0.
\end{align*}
An outbreak occurs if the rate of change of infectious individuals is positive, $\frac{dY}{dt} > 0$, or equivalently,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \implies -\frac{dX}{dt} - \frac{dZ}{dt} &> 0 \\
  \implies \frac{\beta X Y}{N}  - \gamma Y &> 0 ,\\
\implies  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0\\
\implies   \frac{\beta}{\gamma} &> \frac{N}{X}.
\end{align*}
That is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of initially susceptible individuals is large compared to the total population, $\frac{X}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align*}
  \rr \overset{def}{=} \frac{\beta}{\gamma}.
  \end{align*}
  To incorporate randomness into the model, we add noise, namely,
  \begin{align}\label{eq:sir-noise}
    \hat{X}(t) &= X(t) + \epsilon_{X,t}\\
    \hat{Y}(t) &=  N - \hat{X}(t) -\hat{Y}(t)  \nonumber\\
    \hat{Z}(t) &= Z(t) - \epsilon_{Z,t}. \nonumber
  \end{align}
The ``hats'' in Equation \ref{eq:sir-noise} are used to distinguish real, observed data from the ``true'' model without hats.  We are assuming the observations are generated based on the  deterministic ODEs (possibly converted to discrete time) presented in Equation \ref{eq:sir} with time and compartment dependent noise $\epsilon_{X,t}$ and $\epsilon_{Z,t}$.  Since, $N$, the total population, is constant, then $\hat{Y}$ is adjusted accordingly.

In summary, when we discuss estimators for $\rr$ for the SIR model, we mean to say we are forming an estimator of $\rr$ from the given set of data in Eq. \ref{eq:sir-noise},
\begin{align*}
  \textnormal{Data} &= \left \{\left (\hat{X}(t), \hat{Y}(t), \hat{Z}(t) \right ) : t=0, 1, \dots, T\right \}, \\
  \hat{\rr} &= m(\textnormal{Data}),
\end{align*}
where $m$ is a function of the data.

Good properties of an estimator may include (1) being a consistent estimator
\begin{align*}
  \hat{\rr} \overset{P}{\to} \rr \textnormal{ as } T\to \infty,
\end{align*}
where `$\overset{P}{\to}$' stands for convergence in probability or (2) converging to a known probability distribution (convergence in distribution `$\overset{d}{\to}$' \citep{wasserman2004}),
\begin{align*}
\hat{\rr} \overset{d}{\to} F.  
\end{align*}
As we are interested in the hypothesis
\begin{align*}
  H_0:\;& \rr < 1 \\
  H_A:\;& \rr \ge 1
\end{align*}
Then for some $\alpha \in [0,1]$, we would like to know for some $\hat{a}, \hat{b} >0$
\begin{align*}
P(\hat{a} \le \rr \le \hat{b}) \le 1 - \alpha.
\end{align*}
This property is particularly desirable, as we can estimate $\hat{a}$ and $\hat{b}$ from the distribution $F$.  At any rate, we would like to be able to estimate $\hat{a}$ and $\hat{b}$ for a given $\alpha$-level.  We could also make similar hypothesis tests such as  $\rr$ for H1N1 influenza is greater than the reported $\rr$ for seasonal influenza.

@@

\begin{align*}
  \frac{dX}{dt} &= - \frac{\beta_{XEYZ} XY}{N} \\
  \frac{dE}{dt} &= \frac{\beta_{XEYZ} XY}{N}  - \mu_{XEYZ} E\\
  \frac{dY}{dt} &= \mu_{XEYZ} E - \gamma_{XEYZ} Y \\
  \frac{dZ}{dt} &= \gamma_{XEYZ} Y,
\end{align*}
again, with a constant population of $N$ and with known initial values $(X(0), E(0), Y(0), Z(t))$.  The parameter $\mu_{XEYZ}$ is the rate at which latently exposed individuals become infectious.  The subscript ``XEYZ'' is used to emphasize that the disease parameters (especially $\beta$ and $\gamma$) are different than of those in the XYZ model.  The reproduction number for the XEYZ model, is $\rr = \frac{\beta_{XEYZ}}{\gamma_{XEYZ}}$, compared to the reproduction number for the XYZ model, $\rr = \frac{\beta_{XYZ}}{\gamma_{XYZ}}$.

Consider the case when $N=10000, (\beta, \gamma) = (.06, .03)$, $(X(0), E(0), Y(0), Z(t)) = (9500, 0, 500, 0)$ for times $t=1, \dots, 365$.  Let the parameter $\mu$ range between [0.005, 0.250].  This generates data set
$$D_{\mu, \textnormal{XEYZ}} = \left \{(\hat{X}_{XEYZ}(t), \hat{E}_{XEYZ}(t), \hat{Y}_{XEYZ}(t), \hat{Z}_{XEYZ}(t): t=1, \dots, 365\right \},$$
which is observed by researcher 1.  Now assume that researcher 2 misspecifies the model for this data as a SIR (XYZ) model rather than a SEIR (XEYZ) model.  That is, she observes all non-infectious individuals as susceptibles and generates data sets
$$D_{\mu, XYZ} = \left \{\hat{X}_{XYZ}(t) = X_{XEYZ} + E_{XEYZ}, \hat{Y}_{XYZ} = Y_{XEYZ}, \hat{Z}_{XEYZ} = Z_{XYZ} : t = 1, \dots, 365 \right \}.$$
This situation is illustrated in Figure \ref{fig:sir-vs-seir} with $\mu = 0.01$.

@@

Both researchers want to estimate $\rr$ and so fit models to their data sets.  The details of how the models are fit are not so important at the moment (but will be in the later sections), but to fairly compare one model to another assume both researchers use a least squares estimates to estimate the disease parameters and then use the ratio estimator (RE) as an estimate of $\rr$.  That is researcher 1 estimates $\rr$ with
\begin{align*}
  (\hat{\beta}_{XEYZ}, \hat{\gamma}_{XEYZ}, \hat{\mu}_{XEYZ} )&=\arg \min_{\beta, \gamma, \mu} \sum_{t} \left [ \left (X_{XEYZ}(t) - X(t)\right )^2  \right . \\
                                                                 &+\left ( E_{XEYZ}(t) - E(t) \right )^2  \\
  &\left . + \left ( Z_{XEYZ}(t) - Z(t) \right )^2\right ]
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XEYZ}= \frac{\hat{\beta}_{XEYZ}}{\hat{\gamma}_{XEYZ}}.
\end{align*}
Researcher 2 estimates $\rr$ with 
\begin{align*}
(\hat{\beta}_{XYZ}, \hat{\gamma}_{XYZ} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (X_{XYZ}(t) - X(t)\right )^2 + \left ( Z_{XYZ}(t) - Z(t) \right )^2 \right ].
\end{align*}
Then the RE for $\rr^{XYZ}$ is given by Equation \ref{eq:sirls},
\begin{align*}
  \hat{\rr}^{XYZ}= \frac{\hat{\beta}_{XYZ}}{\hat{\gamma}_{XYZ}}.
\end{align*}

We plot the two estimates of $\rr$ as a function of $\mu$ in Figure \ref{fig:r0-alpha}.  We should note we also estimated the standard errors of the estimates of $\rr$, but they do not show up on the figure because they are of the order of $10^{-4}$.  Unsurprisingly, under the SEIR model speficiation, $\hat{\rr}=2$ because there is no noise in the data, and the data is generated from an SEIR model.  However, as we see in Figure \ref{fig:r0-alpha} the RE of $\hat{\rr}$ from the SIR model for small values of $\mu$ (a long latency period) can result in double the size of $\hat{\rr}$!

@@

For those readers not convinced by our small example, we offer this mathematical intuition of the next generation model (NGM).

 The NGM is a generalization of any compartment model at the infection-free steady state ($Y(t)=0$ and all derivatives with respect to time sum to zero). Originally, introduced by \citep{diekmann1990}, \cite{diekmann2009} posit a recipe to find $\rr$ for a wide class of compartment models, including SIR and SEIR. First, define the infected subsystem as ``the equations of the ODE system that describe the production of new infections and changes in state among infected individuals.''  Let $x = (C_1, C_2, \dots, C_m)^T$ where $C_i$ are the different compartments of the infected subsystem.  The steps to find $\rr$ are as follows:

\begin{enumerate}
\item Take the subsystem of infected individuals (including those who have been exposed) and linearize the subsystem about the infection-free steady state
\item Decompose the linearized infected subsystem into $(T + \Sigma )x$ where $T$ is a $m\times m$ matrix of \textit{transmissions} and $\Sigma$ is a $m \times m$ matrix of \textit{transitions}.
\item $\rr$ is the spectral radius (i.e. the dominant eigenvalue) of $-T \Sigma^{-1}$.  
\end{enumerate}

@@

The NGM shows that $\rr$ is the spectral radius of a matrix which is composed of the transmission of disease to susceptible individuals and transition of individuals among the other compartments.  $\rr$ is thus a property of the model.

Following the examples presented in \cite{diekmann1990}, for the SIR (XYZ) model, the linearized infection-free sub-system occurs when $Y(t)=0$ and $X(t)=N$ and is
\begin{align*}
\frac{dY}{dt} &= - \beta_{XYZ} Y + \gamma_{XYZ} .
  \end{align*}
  Then $T= \beta_{XYZ}$ (the rate of new infections) and $\Sigma = -\gamma_{XYZ}$.  Then $K = \frac{\beta_{XYZ}}{\gamma_{XYZ}}$ which is also, trivially, the dominant eigenvalue, and hence, $R_0 = \frac{\beta_{XYZ}}{\gamma_{XYZ}}$.

We can also derive the NGM for the SEIR (XEYZ) model.  We have an infection-free steady state when $E=Y=0$ and set $X=N$.  Then $x = (E, Y)^T$ is the infected subsystem.  The linearized subsystem is
  \begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dE}{dt} &= \beta_{XEYZ} Y  - \mu_{XEYZ} E\vspace{.5em}\\
    \frac{dY}{dt} &=  \mu_{XEYZ} E  - \gamma_{XEYZ} Y \\
  \end{array}
  \right .
  \end{align*}
  Then $T = \left ( \begin{array}{cc} 0 & \beta_{XEYZ} \\ 0 & 0  \end{array} \right )$ and $\Sigma = \left ( \begin{array}{cc} -\mu_{XEYZ} & 0 \\ \mu_{XEYZ} & - \gamma_{XEYZ} \end{array} \right )$, and $K =\left ( \begin{array}{cc} 0& \beta_{XEYZ} / \gamma_{XEYZ} \\ 0 & 0 \end{array} \right  )$.

  Then the spectral radius of $K$ is $R_0  = \frac{\beta_{XEYZ}}{\gamma_{XEYZ}}$, which makes sense as exposed individuals are not contributing to the number of new infections.

  Although the $\rr$ ends up being similar (infection rate over the recovery rate) for both the SIR (XYZ) and SEIR (XEYZ) models, the derivations are different because of the addition of the exposed compartment.  In general, more complicated models will result in different derivations of $T$ and $\Sigma$ and hence different derivations of $\rr$.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\arg \min_{\beta, \gamma} \sum_{t} \left [ \left (\hat{X}(t) - X(t)\right )^2 + \left ( \hat{Z}(t) - Z(t) \right )^2 \right ]
\end{align*}
Then the ratio estimator (RE) estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{\rr}= \frac{\hat{\beta}}{\hat{\gamma}}.
\end{align}

The estimate resulting from this method are often found via a grid search of $\beta$ and $\gamma$ values to minimize the squared error.  The estimates can also be found using more sophisticated optimization algorithms. Since we typically cannot explicitly write down the partial derivatives of $S$, $I$, and $R$ with respect to $\beta$ and $\gamma$, we can only use minimization algorithms which do not rely on an explicit gradient such as Nelder-Mead simplex minimization \citep{nelder-mead1965}.

A reasonable question one may ask is why use the $L_2$-norm as opposed to $L_1$ or some other similarity score.  A possible answer is that if we were to assume Gaussian noise, then the $L_2$-norm would be equivalent to the maximum likelihood estimation.  Another answer is that $L_2$ is a continuously differentiable function and hence is easier to compute, which allows for sensitivity analysis to  be conducted more easily.  Finally, the RE estimate can be found without writing down explicit assumptions about the noise within the model.  However, we cannot guarantee properties of consistency or convergence to a known distribution without more explicit assumptions.
